---
title: "Project1"
author: "Ryan Friedman"
format: html
editor: visual
---

```{r}
#| include: false
# load packages that may be used
library(tidyverse)
library(readxl)
library(DBI)
library(knitr)
library(httr)
library(jsonlite)
library(psych)
library(gt)
library(tidycensus)
```

```{r}
#| include: false
APIKey <- "5a644b1ecc301cab1403e0940017740912735ef9" # R.Friedman API Key for census.gov
```

## Test Query of API

To start, we'll do a test query of the API and review the format of the data

```{r}
TestData <- GET(paste0("https://api.census.gov/data/2022/acs/acs1/pums?get=SEX,PWGTP,AGEP&for=state:06&SCHL=24&key=", APIKey))
```

```{r}
TestParsed <- fromJSON(rawToChar(TestData$content))
head(TestParsed)
```

Looks like we get a matrix of all characters. Also the first row contains the variable names. We need to:

1.  Set the names based on the first row

2.  Delete the first row

3.  Convert to tibble

4.  Convert variables into numeric or factors

```{r}
TestCleaned <- TestParsed
colnames(TestCleaned) <- TestCleaned[1, ] # set the column names from the first row
TestCleaned <- as_tibble(TestCleaned[-1, ]) # delete first row and convert to tibble
TestCleaned <- TestCleaned |>
  mutate(SEX = factor(SEX), PWGTP = as.numeric(PWGTP), AGEP = as.numeric(AGEP), SCHL = factor(SCHL), state = factor(state)) # setting variables as numeric or factor
head(TestCleaned)
str(TestCleaned)
```

## Helper Function `IntoTibble()`

`IntoTibble()` function is designed to take the census data returned from the `GET()` and convert into a tibble similar to the steps above

```{r}
IntoTibble <- function (CensusGet){ # input is the get() data from the census API
  CensusParsed <- fromJSON(rawToChar(CensusGet$content)) # extracts the content from the API get
  colnames(CensusParsed) <- CensusParsed[1, ] # sets variable names based on the first row
  CensusParsed <- as_tibble(CensusParsed[-1, ]) # deletes first row and converts to tibble
  CensusParsed
}

head(IntoTibble(TestData)) # testing against the above data
```

Test of `IntoTibble()` function is successful.

## Time Range String Conversion Function `StringToTime()`

JWAP and JWDP both associate time codes with character strings representing a time range. `StringToTime()` will convert the character string into the middle value of the time range as minutes past midnight (i.e. 1:15am would be 75 minutes past midnight)

Example character strings: "9:55 p.m. to 9:59 p.m." "11:00 p.m. to 11:04 p.m." "4:10 a.m. to 4:14 a.m."

If we split the string by both " " and ":", we'll get a vector of 7 predictable elements:

1.  Hours for start time

2.  Minutes for start time

3.  "a.m." or "p.m."

4.  "to" (not used)

5.  Hours for end time

6.  Minutes for end time

7.  "a.m." or "p.m."

Writing and testing the `StringToTime()` function:

```{r}
StringToTime <- function (string) {
  vector <- unlist(strsplit(string, "[ :]")) # split the string by space and colon as described above, then unlist() resulting in a vector
  if(vector[1] == "12") {vector[1] <- "0"} # 12am counts as zero hours past midnight, so 12 should become 0
  if(vector[5] == "12") {vector[5] <- "0"} 
  StartTime <- as.numeric(vector[1]) * 60 + as.numeric(vector[2]) # calculates time as hours * 60 + minutes
  if(vector[3] == "p.m.") {StartTime <- StartTime + 12 * 60} # adds 12*60 minutes if p.m.
  EndTime <- as.numeric(vector[5]) * 60 + as.numeric(vector[6])
  if(vector[7] == "p.m.") {EndTime <- EndTime + 12 * 60}
  MeanTime <- mean(c(StartTime, EndTime)) # calculates the center of the range as the mean between start and end
  return(MeanTime) # returns middle of the range as minutes past midnight
}

StringToTime("9:55 p.m. to 9:59 p.m.") # testing example strings
StringToTime("11:00 p.m. to 11:04 p.m.")
StringToTime("4:10 a.m. to 4:14 a.m.")
```

Test of `StringToTime()` is successful.

## `JWAPTibble`

JWAP is a numeric code that maps onto an irregular range of times. We will create `JWAPTibble` to correlate the time code to the middle value of the time range as minutes past midnight (i.e. 1:15am would be 75 minutes past midnight)

```{r}
JWAP <- fromJSON("Project1DataFiles/JWAP.json") # this JWAP descriptor is from https://api.census.gov/data/2022/acs/acs1/pums/variables/JWAP.json
JWAP <- unlist(JWAP$values$item) # gets to the useful content of the file
str(JWAP)
head(JWAP)
```

Now we have a named vector of 286 character strings. The names are per the various JWAP codes. The strings are the descriptions of the time interval. Lets convert into a tibble and apply `StringToTime()`.

```{r}
JWAPTibble <- as_tibble(names(JWAP)) # begin with just the names of JWAP, which are the codes
JWAPTibble <- mutate(JWAPTibble, string = JWAP) # add a column of the string content
JWAPTibble <- mutate(JWAPTibble, time = sapply(string, StringToTime)) # use StringToTime() to make the third column of time values
JWAPTibble <- mutate(JWAPTibble, value = as.numeric(value)) # convert value to numeric
head(JWAPTibble)
```

JWAPTibble is now ready to help us map JWAP codes to a proper time value.

## `JWDPTibble`

Unfortunately JWDP has different time codes and ranges than JWAP, so the above process will need to be repeated to create `JWDPTibble`

```{r}
JWDP <- fromJSON("Project1DataFiles/JWDP.json") # this JWDP descriptor is from https://api.census.gov/data/2022/acs/acs1/pums/variables/JWDP.json
JWDP <- unlist(JWDP$values$item) # gets to the useful content of the file

JWDPTibble <- as_tibble(names(JWDP)) # begin with just the names of JWDP, which are the codes
JWDPTibble <- mutate(JWDPTibble, string = JWDP) # add a column of the string content
JWDPTibble <- mutate(JWDPTibble, time = sapply(string, StringToTime)) # use StringToTime() to make the third column of time values
JWDPTibble <- mutate(JWDPTibble, value = as.numeric(value)) # convert value to numeric
head(JWDPTibble)
```

JWDPTibble is now ready to help us map JWDP codes to a proper time value.

## Single Year Query Function `QCensus()`

`QCensus()` will query the API and allow user to specify:

- Year of survey

- Numeric variables (AGEP, GASP, GRPIP, JWAP, JWDP, JWMNP)

  - AGEP and PWGTP are default

  - PWGTP is always returned

- Categorical variables (FER, HHL, SCH, SCHL, SEX)

  - SEX is default

- Geography level (region, division, state)

- Optional geography subsetting within the API call

```{r}
QCensus <- function (year = 2022,
                     numVars = c("AGEP"), 
                     catVars = c("SEX"), 
                     geoLevel = "state",
                     geoSubset = "56", # setting Wyoming as default state
                     APIKey = "5a644b1ecc301cab1403e0940017740912735ef9") # using R.Friedman API key as default
  { 
  
  if (geoSubset == "all") {geoSubset = "*"} # converts entry of "all" into "*" 
  
  if (!is.numeric(year)){stop("Year must be a number")} # first checks if year is a number
  if (year == 2020 | year %% 1 != 0 | year > 2022 | year < 2010){ # check if year is valid integer between 2010 and 2022, excluding 2020
    stop("Invalid year.  Must be an integer from 2010 to 2022, excluding 2020")}
  
  ValidNumVars <- c("PWGTP", "AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP") # list of all valid variables for numVars
  ValidCatVars <- c("FER", "HHL", "SCH", "SCHL", "SEX") # list of all valid variables for catVars
  ValidLevelVars <- c("region", "division", "state") # list of all valid variables for geoLevel
  
  if (!all(numVars %in% ValidNumVars)) { # confirm numeric variables are valid
    stop("Invalid numeric variable(s).  Valid numeric variables are: PWGTP, AGEP, GASP, GRPIP, JWAP, JWDP, JWMNP.")}  
  if (!all(catVars %in% ValidCatVars)) { # confirm categorical variables are valid
    stop("Invalid categorical variable(s).  Valid categorical variables are: FER, HHL, SCH, SCHL, SEX.")}    
  if (!all(geoLevel %in% ValidLevelVars)) { # confirm geography level variables are valid
    stop("Invalid geography level variable(s).  Valid geography levels are: region, division, state.")}    
  
  if (length(unique(numVars)) != length(numVars)) { # checking for duplicated numeric variables
    stop("Error: Duplicated numeric variable")}
  if (length(unique(catVars)) != length(catVars))  { # checking for duplicated categorical variables
    stop("Error: Duplicated categorical variable")}
  
  numVars <- numVars[numVars != "PWGTP"] # pulling out PWGTP from numVars since that variable is always included anyways
  
  if(length(numVars) < 1 ) { # checks minimum input requirement for numVars
    stop("Must specify at least one numeric variable other than PWGTP")}  
  if(length(catVars) < 1 ) { # checks minimum input requirement for catVars
    stop("Must specify at least one categorical variable")} 
  
  # establish the valid codes for regions, divisions, and states
  ValidRegions <- c("1", "2", "3", "4", "*")
  ValidDivisions <- c("1", "2", "3", "4", "5", "6", "7", "8", "9", "*")
  ValidStates <- c("01", "02", "04", "05", "06", "08", "09", "10", "11", "12", "13", "15", "16", "17", "18", "19", "20", 
                   "21", "22", "23", "24", "25", "26", "27", "28", "29", "30", "31", "32", "33", "34", "35", "36", "37", 
                   "38", "39", "40", "41", "42", "44", "45", "46", "47", "48", "49", "50", "51", "53", "54", "55", "56", "*")
  
  # add corresponding names (to be used later).  "All" is simply a placeholder that will not be used
  names(ValidDivisions) <- c("New England", "Middle Atlantic", "East North Central", "West North Central", 
                             "South Atlantic", "East South Central", "West South Central", "Mountain", "Pacific", "All")
  names(ValidRegions) <- c("Northeast", "Midwest", "South", "West", "All")
  names(ValidStates) <- c("Alabama", "Alaska", "Arizona", "Arkansas", "California", "Colorado", "Connecticut", "Delaware", 
                          "District of Columbia", "Florida", "Georgia", "Hawaii", "Idaho", "Illinois", "Indiana", "Iowa", 
                          "Kansas", "Kentucky", "Louisiana", "Maine", "Maryland", "Massachusetts", "Michigan", "Minnesota", 
                          "Mississippi", "Missouri", "Montana", "Nebraska", "Nevada", "New Hampshire", "New Jersey", 
                          "New Mexico", "New York", "North Carolina", "North Dakota", "Ohio", "Oklahoma", "Oregon", 
                          "Pennsylvania", "Rhode Island", "South Carolina", "South Dakota", "Tennessee", "Texas", "Utah", 
                          "Vermont", "Virginia", "Washington", "West Virginia", "Wisconsin", "Wyoming", "All")
  
  if (geoLevel == "region" & !(geoSubset %in% ValidRegions)) { # check if geoSubset matches valid regions
    stop("Invalid region code")} 
  if (geoLevel == "division" & !(geoSubset %in% ValidDivisions)) { # check if geoSubset matches valid divisions
    stop("Invalid division code")} 
  if (geoLevel == "state" & !(geoSubset %in% ValidStates)) { # check if geoSubset matches valid states
    stop("Invalid state code")} 
  
  APIURL <- paste0("https://api.census.gov/data/", year, "/acs/acs1/pums?get=PWGTP,") # begins the URL including year and PWGTP
  APIURL <- paste0(APIURL, paste(numVars, collapse = ","), ",", paste(catVars, collapse = ",")) # adds numeric variables and categorical variables
  APIURL <- paste0(APIURL, "&for=", geoLevel, ":", geoSubset) # adds geoLevel and subset
  APIURL <- paste0(APIURL, "&key=", APIKey) # adds API key
  
  QCensusParsed <- IntoTibble(GET(APIURL)) # Gets tibble from URL
  
  for (i in c("PWGTP", numVars)) { # cycle through all relevant numeric variables, adding PWGTP back in
    if (i %in% c("PWGTP", "AGEP", "GASP", "GRPIP", "JWMNP")) { # these variables are pure numbers
      QCensusParsed[[i]] <- as.numeric(QCensusParsed[[i]]) # convert to numeric
    }
    if (i == "JWAP") { # converting JWAP times
      QCensusParsed <- QCensusParsed |>
        mutate(JWAP = as.numeric(JWAP)) |> # convert JWAP to numeric to match JWAPTibble
        left_join(JWAPTibble, by = c("JWAP" = "value")) |> # join with JWAPTibble, keeping the original QCensusParsed and correlating the codes
        mutate(JWAP = time) |> # we can now overwrite JWAP in QCensusParsed with the proper time values
        select(!c(string, time)) # remove unused string and time variables
    }
    if (i == "JWDP") { #converting JWDP times, same as above
      QCensusParsed <- QCensusParsed |>
        mutate(JWDP = as.numeric(JWDP)) |>
        left_join(JWDPTibble, by = c("JWDP" = "value")) |>
        mutate(JWDP = time) |>
        select(!c(string, time))      
    }   
  }
  
  for (i in catVars){ # cycle through all relevant categorical variables in catVars
    QCensusParsed[[i]] <- factor(QCensusParsed[[i]]) # convert to factor
  } 
  
  QCensusParsed[[geoLevel]] <- factor(QCensusParsed[[geoLevel]]) # convert appropriate geoLevel to a factor

  # Consider setting factor labels for catVar and/or geoSubset levels if time allows. 
  
  return(QCensusParsed)
}
```

Let's test some uses of QCensus()

```{r}
system.time(print(QCensus())) # running defaults
```

```{r}
system.time(print(QCensus(year = 2021, numVars = c("JWAP", "JWDP"), catVars = c("SCHL", "SEX"), geoLevel = "state", geoSubset = "17" ))) # adding some time and categorical variables
```

```{r}
system.time(print(QCensus(year = 2018, numVars = c("AGEP", "JWDP", "JWAP", "GRPIP"), catVars = c("FER"), geoLevel = "division", geoSubset = "4" ))) # trying a division and some other variables
```

```{r}
# system.time(BigTest <- QCensus(year = 2015, numVars = c("PWGTP", "AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP"), catVars = c("FER", "HHL", "SCH", "SCHL", "SEX"), geoLevel = "state", geoSubset = "all"))

# BigTest took ~6 minutes and returned ~3.1 million rows.  Compute time ("user") was 41 sec, so the code is not the bottleneck.  Not going to have this running in the .qmd, but it did work!
```

Runtime is manageable but increases with the size of the query (per geoLevel and geoSubset).  Majority of runtime is waiting on the API rather than compute, so code is reasonably optimized.  Be wary of setting geoSubset = "all" resulting in runtimes over 5 min. 

All tests of `QCensus()` are successful.  

## Multi-Year Query Function `QMultiCensus()`
