---
title: "Project1"
author: "Ryan Friedman"
format: html
editor: visual
---

```{r}
#| include: false
# load packages that may be used
library(tidyverse)
library(readxl)
library(DBI)
library(knitr)
library(httr)
library(jsonlite)
library(psych)
library(gt)
library(tidycensus)
```

```{r}
#| include: false
APIKey <- "5a644b1ecc301cab1403e0940017740912735ef9" # R.Friedman API Key for census.gov
```

## Test Query of API

To start, we'll do a test query of the API and review the format of the data

```{r}
TestData <- GET(paste0("https://api.census.gov/data/2022/acs/acs1/pums?get=SEX,PWGTP,AGEP&for=state:06&SCHL=24&key=", APIKey))
```

```{r}
TestParsed <- fromJSON(rawToChar(TestData$content))
head(TestParsed)
```

Looks like we get a matrix of all characters. Also the first row contains the variable names. We need to:

1.  Set the names based on the first row

2.  Delete the first row

3.  Convert to tibble

4.  Convert variables into numeric or factors

```{r}
TestCleaned <- TestParsed
colnames(TestCleaned) <- TestCleaned[1, ] # set the column names from the first row
TestCleaned <- as_tibble(TestCleaned[-1, ]) # delete first row and convert to tibble
TestCleaned <- TestCleaned |>
  mutate(SEX = factor(SEX), PWGTP = as.numeric(PWGTP), AGEP = as.numeric(AGEP), SCHL = factor(SCHL), state = factor(state)) # setting variables as numeric or factor
head(TestCleaned)
str(TestCleaned)
```

## Helper Function `IntoTibble()`

`IntoTibble()` function is designed to take the census data returned from the `GET()` and convert into a tibble similar to the steps above

```{r}
IntoTibble <- function (CensusGet){ # input is the get() data from the census API
  CensusParsed <- fromJSON(rawToChar(CensusGet$content)) # extracts the content from the API get
  colnames(CensusParsed) <- CensusParsed[1, ] # sets variable names based on the first row
  CensusParsed <- as_tibble(CensusParsed[-1, ]) # deletes first row and converts to tibble
  CensusParsed
}

head(IntoTibble(TestData)) # testing against the above data
```

Test of `IntoTibble()` function is successful.

## Single Year Query Function `QCensus()`

`QCensus()` will query the API and allow user to specify:

- Year of survey

- Numeric variables (AGEP, GASP, GRPIP, JWAP, JWDP, JWMNP)

  - AGEP and PWGTP are default

  - PWGTP is always returned

- Categorical variables (FER, HHL, SCH, SCHL, SEX)

  - SEX is default

- Geography level (region, division, state)

- Optional geography subsetting within the API call

```{r}
QCensus <- function (year = 2022,
                     numVars = "AGEP",
                     catVars = "SEX", 
                     geoRegion = "state",
                     geoSubset = "06"){
  
}
```

## Multi-Year Query Function `QMultiCensus()`
