---
title: "Project1"
author: "Ryan Friedman"
format: html
editor: visual
---

```{r}
#| include: false
# load packages that may be used
library(tidyverse)
library(readxl)
library(DBI)
library(knitr)
library(httr)
library(jsonlite)
library(psych)
library(gt)
library(tidycensus)
library(quantreg)
```

```{r}
#| include: false
APIKey <- "5a644b1ecc301cab1403e0940017740912735ef9" # R.Friedman API Key for census.gov
```

## 1. Test Query of API

To start, we'll do a test query of the API and review the format of the PUMS Census data

```{r}
TestData <- GET(paste0("https://api.census.gov/data/2022/acs/acs1/pums?get=SEX,PWGTP,AGEP&for=state:06&SCHL=24&key=", APIKey))
```

```{r}
TestParsed <- fromJSON(rawToChar(TestData$content))
head(TestParsed)
```

Looks like we get a data frame of all characters. Also the first row contains the variable names. We need to:

1.  Set the names based on the first row

2.  Delete the first row

3.  Convert to tibble

4.  Convert variables into numeric or factors

```{r}
TestCleaned <- TestParsed
colnames(TestCleaned) <- TestCleaned[1, ] # set the column names from the first row
TestCleaned <- as_tibble(TestCleaned[-1, ]) # delete first row and convert to tibble
TestCleaned <- TestCleaned |>
  mutate(SEX = factor(SEX), PWGTP = as.numeric(PWGTP), AGEP = as.numeric(AGEP), SCHL = factor(SCHL), state = factor(state)) # setting variables as numeric or factor
head(TestCleaned)
str(TestCleaned)
```

Looks good. We'll work on automating this process later.

## 2. Helper Function `IntoTibble()`

`IntoTibble()` function is designed to take the census data returned from the `GET()` and convert into a tibble similar to the steps above

```{r}
IntoTibble <- function (CensusGet){ # input is the get() data from the census API
  CensusParsed <- fromJSON(rawToChar(CensusGet$content)) # extracts the content from the API get
  colnames(CensusParsed) <- CensusParsed[1, ] # sets variable names based on the first row
  CensusParsed <- as_tibble(CensusParsed[-1, ]) # deletes first row and converts to tibble
  CensusParsed
}

head(IntoTibble(TestData)) # testing against the above data
```

Test of `IntoTibble()` function is successful.

## 3. Time Range String Conversion Function `StringToTime()`

JWAP and JWDP both associate time codes with character strings representing a time range. `StringToTime()` will convert the character string into the middle value of the time range as minutes past midnight (i.e. 1:15am would be 75 minutes past midnight)

Example character strings: "9:55 p.m. to 9:59 p.m." "11:00 p.m. to 11:04 p.m." "4:10 a.m. to 4:14 a.m."

If we split the string by both " " and ":", we'll get a vector of 7 predictable elements:

1.  Hours for start time

2.  Minutes for start time

3.  "a.m." or "p.m."

4.  "to" (not used)

5.  Hours for end time

6.  Minutes for end time

7.  "a.m." or "p.m."

Testing the `StringToTime()` function

```{r}
StringToTime <- function (string) {
  if(string == "N/A (not a worker; worker who worked from home)") {return(NA)} # the string associated with code 0 will just return NA and skip the parsing
  vector <- unlist(strsplit(string, "[ :]")) # split the string by space and colon as described above, then unlist() resulting in a vector
  if(vector[1] == "12") {vector[1] <- "0"} # 12am counts as zero hours past midnight, so 12 should become 0
  if(vector[5] == "12") {vector[5] <- "0"} 
  StartTime <- as.numeric(vector[1]) * 60 + as.numeric(vector[2]) # calculates time as hours * 60 + minutes
  if(vector[3] == "p.m.") {StartTime <- StartTime + 12 * 60} # adds 12*60 minutes if p.m.
  EndTime <- as.numeric(vector[5]) * 60 + as.numeric(vector[6])
  if(vector[7] == "p.m.") {EndTime <- EndTime + 12 * 60}
  MeanTime <- mean(c(StartTime, EndTime)) # calculates the center of the range as the mean between start and end
  return(MeanTime) # returns middle of the range as minutes past midnight
}
```

Testing the `StringToTime()` function

```{r}
StringToTime("9:55 p.m. to 9:59 p.m.") 
StringToTime("11:00 p.m. to 11:04 p.m.")
StringToTime("4:10 a.m. to 4:14 a.m.")
```

Test of `StringToTime()` is successful.

## 4. `JWAPTibble`

JWAP is a numeric code that maps onto an irregular range of times. We will create `JWAPTibble` to correlate the time code to the middle value of the time range as minutes past midnight (i.e. 1:15am would be 75 minutes past midnight)

```{r}
JWAP <- fromJSON("Project1DataFiles/JWAP.json") # this JWAP descriptor is from https://api.census.gov/data/2022/acs/acs1/pums/variables/JWAP.json
JWAP <- unlist(JWAP$values$item) # gets to the useful content of the file
str(JWAP)
head(JWAP)
```

Now we have a named vector of 286 character strings. The names are per the various JWAP codes. The strings are the descriptions of the time interval. Lets convert into a tibble and apply `StringToTime()`.

```{r}
JWAPTibble <- as_tibble(names(JWAP)) # begin with just the names of JWAP, which are the codes
JWAPTibble <- mutate(JWAPTibble, string = JWAP) # add a column of the string content
JWAPTibble <- mutate(JWAPTibble, time = sapply(string, StringToTime)) # use StringToTime() to make the third column of time values
JWAPTibble <- mutate(JWAPTibble, value = as.numeric(value)) # convert value to numeric
JWAPTibble
```

`JWAPTibble` is now ready to help us map JWAP codes to a proper time value.

## 5. `JWDPTibble`

Unfortunately JWDP has different time codes and ranges than JWAP, so the above process will need to be repeated to create `JWDPTibble`

```{r}
JWDP <- fromJSON("Project1DataFiles/JWDP.json") # this JWDP descriptor is from https://api.census.gov/data/2022/acs/acs1/pums/variables/JWDP.json
JWDP <- unlist(JWDP$values$item) # gets to the useful content of the file

JWDPTibble <- as_tibble(names(JWDP)) # begin with just the names of JWDP, which are the codes
JWDPTibble <- mutate(JWDPTibble, string = JWDP) # add a column of the string content
JWDPTibble <- mutate(JWDPTibble, time = sapply(string, StringToTime)) # use StringToTime() to make the third column of time values
JWDPTibble <- mutate(JWDPTibble, value = as.numeric(value)) # convert value to numeric
JWDPTibble
```

`JWDPTibble` is now ready to help us map JWDP codes to a proper time value.

## 6. Single Year Query Function `QCensus()`

`QCensus()` will query the API and allow user to specify:

- Year of survey

- Numeric variables (AGEP, GASP, GRPIP, JWAP, JWDP, JWMNP)

  - AGEP and PWGTP are default

  - PWGTP is always returned

- Categorical variables (FER, HHL, SCH, SCHL, SEX)

  - SEX is default

- Geography level (region, division, state)

- Optional geography subsetting within the API call

```{r}
QCensus <- function (year = 2022,
                     numVars = c("AGEP"), # default numV
                     catVars = c("SEX"), 
                     geoLevel = "state",
                     geoSubset = "56", # setting Wyoming as default state
                     APIKey = "5a644b1ecc301cab1403e0940017740912735ef9") # using R.Friedman API key as default
  { 
  
  if (geoSubset == "all") {geoSubset = "*"} # converts entry of "all" into "*" 
  
  if (!is.numeric(year)){stop("Year must be a number")} # first checks if year is a number
  if (year == 2020 | year %% 1 != 0 | year > 2022 | year < 2010){ # check if year is valid integer between 2010 and 2022, excluding 2020
    stop("Invalid year.  Must be an integer from 2010 to 2022, excluding 2020")}
  
  ValidNumVars <- c("PWGTP", "AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP") # list of all valid variables for numVars
  ValidCatVars <- c("FER", "HHL", "SCH", "SCHL", "SEX") # list of all valid variables for catVars
  ValidLevelVars <- c("region", "division", "state") # list of all valid variables for geoLevel
  
  if (!all(numVars %in% ValidNumVars)) { # confirm numeric variables are valid
    stop("Invalid numeric variable(s).  Valid numeric variables are: PWGTP, AGEP, GASP, GRPIP, JWAP, JWDP, JWMNP.")}  
  if (!all(catVars %in% ValidCatVars)) { # confirm categorical variables are valid
    stop("Invalid categorical variable(s).  Valid categorical variables are: FER, HHL, SCH, SCHL, SEX.")}    
  if (!all(geoLevel %in% ValidLevelVars)) { # confirm geography level variables are valid
    stop("Invalid geography level variable(s).  Valid geography levels are: region, division, state.")}    
  
  if (length(unique(numVars)) != length(numVars)) { # checking for duplicated numeric variables
    stop("Error: Duplicated numeric variable")}
  if (length(unique(catVars)) != length(catVars))  { # checking for duplicated categorical variables
    stop("Error: Duplicated categorical variable")}
  
  numVars <- numVars[numVars != "PWGTP"] # pulling out PWGTP from numVars since that variable is always included anyways
  
  if(length(numVars) < 1 ) { # checks minimum input requirement for numVars
    stop("Must specify at least one numeric variable other than PWGTP")}  
  if(length(catVars) < 1 ) { # checks minimum input requirement for catVars
    stop("Must specify at least one categorical variable")} 
  
  # Establish the valid codes for regions, divisions, and states
  ValidRegions <- c("1", "2", "3", "4", "*")
  ValidDivisions <- c("1", "2", "3", "4", "5", "6", "7", "8", "9", "*")
  ValidStates <- c("01", "02", "04", "05", "06", "08", "09", "10", "11", "12", "13", "15", "16", "17", "18", "19", "20", 
                   "21", "22", "23", "24", "25", "26", "27", "28", "29", "30", "31", "32", "33", "34", "35", "36", "37", 
                   "38", "39", "40", "41", "42", "44", "45", "46", "47", "48", "49", "50", "51", "53", "54", "55", "56", "*")
  
  if (geoLevel == "region" & !(geoSubset %in% ValidRegions)) { # check if geoSubset matches valid regions
    stop("Invalid region code")} 
  if (geoLevel == "division" & !(geoSubset %in% ValidDivisions)) { # check if geoSubset matches valid divisions
    stop("Invalid division code")} 
  if (geoLevel == "state" & !(geoSubset %in% ValidStates)) { # check if geoSubset matches valid states
    stop("Invalid state code")} 
  
  APIURL <- paste0("https://api.census.gov/data/", year, "/acs/acs1/pums?get=PWGTP,") # begins the URL including year and PWGTP
  APIURL <- paste0(APIURL, paste(numVars, collapse = ","), ",", paste(catVars, collapse = ",")) # adds numeric variables and categorical variables
  APIURL <- paste0(APIURL, "&for=", geoLevel, ":", geoSubset) # adds geoLevel and subset
  APIURL <- paste0(APIURL, "&key=", APIKey) # adds API key
  
  QCensusParsed <- IntoTibble(GET(APIURL)) # Gets tibble from URL
  
  for (i in c("PWGTP", numVars)) { # cycle through all relevant numeric variables, adding PWGTP back in
    if (i %in% c("PWGTP", "AGEP", "GASP", "GRPIP", "JWMNP")) { # these variables are pure numbers
      QCensusParsed[[i]] <- as.numeric(QCensusParsed[[i]]) # convert to numeric
    }
    if (i == "JWAP") { # converting JWAP times
      QCensusParsed <- QCensusParsed |>
        mutate(JWAP = as.numeric(JWAP)) |> # convert JWAP to numeric to match JWAPTibble
        left_join(JWAPTibble, by = c("JWAP" = "value")) |> # join with JWAPTibble, keeping the original QCensusParsed and correlating the codes
        mutate(JWAP = time) |> # we can now overwrite JWAP in QCensusParsed with the proper time values
        select(!c(string, time)) # remove unused string and time variables
    }
    if (i == "JWDP") { #converting JWDP times, same as above
      QCensusParsed <- QCensusParsed |>
        mutate(JWDP = as.numeric(JWDP)) |>
        left_join(JWDPTibble, by = c("JWDP" = "value")) |>
        mutate(JWDP = time) |>
        select(!c(string, time))      
    }   
  }
  
  for (i in catVars){ # cycle through all relevant categorical variables in catVars
    # factor levels do not order as desired for plotting later, so we have to force it here
    numLevels <- as.character(sort(as.numeric(unique(QCensusParsed[[i]])))) # create a vector of the categorical factor levels as characters but sorted numerically
    QCensusParsed[[i]] <- factor(QCensusParsed[[i]], levels = numLevels) # convert to factor and apply levels in the correct order per numLevels
  } 
  
  QCensusParsed[[geoLevel]] <- factor(QCensusParsed[[geoLevel]]) # convert appropriate geoLevel to a factor

  class(QCensusParsed) <- c("census", class(QCensusParsed)) # adds class "census" to the tibble output
  
  return(QCensusParsed)
}
```

Let's test some uses of `QCensus()`

First test some error use cases:

```{r}
#| error: true # allows HTML to render despite error
QCensus(year = 2003)
```
```{r}
#| error: true # allows HTML to render despite error
QCensus(numVars = "ERRORINPUT")
```
```{r}
#| error: true # allows HTML to render despite error
QCensus(numVars = "PWGTP")
```
```{r}
#| error: true # allows HTML to render despite error
QCensus(catVars = NULL)
```

Now test some correct use cases:

```{r}
system.time(print(QCensus())) # running defaults
```

```{r}
system.time(print(QCensus(year = 2021, numVars = c("JWAP", "JWDP"), catVars = c("SCHL", "SEX"), geoLevel = "state", geoSubset = "17" ))) # adding some time and categorical variables
```

```{r}
system.time(print(QCensus(year = 2018, numVars = c("AGEP", "JWDP", "JWAP", "JWMNP", "GRPIP"), catVars = c("FER"), geoLevel = "division", geoSubset = "4" ))) # trying a division and some other variables
```

Note: Departure time for work "JWDP" and work arrival time "JWAP" calculations are approximately validated by work travel time "JWMNP". Calculation is not always exact since JWDP and JWAP assume the middle of the time range reported.

```{r}
# system.time(BigTest <- QCensus(year = 2015, numVars = c("PWGTP", "AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP"), catVars = c("FER", "HHL", "SCH", "SCHL", "SEX"), geoLevel = "state", geoSubset = "all"))

# BigTest took ~6 minutes and returned ~3.1 million rows.  Compute time ("user") was 41 sec, so the code is not the bottleneck.  Not going to have this running in the .qmd, but it did work!
```

Runtime is manageable but increases with the size of the query (per geoLevel and geoSubset). Majority of runtime is waiting on the API rather than compute, so code is reasonably optimized. Be wary of setting geoSubset = "all" resulting in runtimes over 5 min.

All tests of `QCensus()` are successful.

## 7. Multi-Year Query Function `QMultiCensus()`

`QMultiCensus()` will build on `QCensus()` to query multiple years of data.

The user will specify a valid range of years (in addition to the other `QCensus()` inputs), and `QMultiCensus()` will return a single tibble with combined data and a new year variable

```{r}
QMultiCensus <- function(CensusYears = c(2010:2011), # 
                         numVars = c("AGEP"), 
                         catVars = c("SEX"), 
                         geoLevel = "state", 
                         geoSubset = "56", # setting Wyoming as default state
                         APIKey = "5a644b1ecc301cab1403e0940017740912735ef9") {# using R.Friedman API key as default)
  
  for(i in CensusYears) {
    if(i %% 1 != 0 | i > 2022 | i < 2010){ # confirms all CensusYears are within the valid range
      stop("Invalid range of years.  Range must be within 2010 to 2022.  2020 is skipped.")}
  }
  CensusYears <- CensusYears[CensusYears != 2020] # removes 2020 from CensusYears, if present
  CensusYears <- unique(CensusYears) # removes any duplicates, if present
  
  QMultiParsed = NULL # need to start with placeholder here in order for bind_rows to work initally
  
  for(i in CensusYears) { # loop through all census years we want
    QMultiParsed <- QCensus(i, numVars, catVars, geoLevel, geoSubset, APIKey) |> # call QCensus() on the specific year and variables desired
      mutate(year = i) |> # add a year column 
      bind_rows(QMultiParsed) # attach resulting tibble to existing tibble (building longer each time)
  }
  
  QMultiParsed <- arrange(QMultiParsed, year) # arranges results in lowest to highest year
  
  return(QMultiParsed)
}
```

Test `QMultiCensus()`

```{r}
system.time(RunTest <- QMultiCensus(CensusYears = c(2018:2021))) # default variables.  testing exclusion of year 2020
str(RunTest) # check structure
unique(RunTest$year) # verify years present in result
remove(RunTest) # delete temp variable
```

```{r}
system.time(RunTest <- QMultiCensus(CensusYears = c(2010:2022), numVars = c("JWAP", "JWDP"), catVars = c("SCHL", "SEX"), geoLevel = "state", geoSubset = "02" )) # Alaska data from 2010 to 2022
str(RunTest) # check structure
unique(RunTest$year) # verify years present in result
remove(RunTest) # delete temp variable
```

All tests of `QMultiCensus()` are successful.

## 8. `summary()` for Census Data

The "census" class has been added to the tibble returned by `QCensus()` above. This also carries through to `QMultiCensus`

We will write our own function to direct how `summary()` is applied to "census" objects

```{r}
summary.census <- function(CensusTibble, 
                           numVars = c("AGEP", "GASP", "GRPIP", "JWMNP", "JWAP"), # by default includes applicable numeric variables.  excludes PWGTP, JWDP, and year
                           ## Note:  JWAP was deliberately included so that it could be analyzed in the Data Investigation section
                           catVars = c("FER", "HHL", "SCH", "SCHL", "SEX", "region", "division", "state")){ # by default includes all possible categorical variables (also geoLevel if the tibble includes multiple SubSets)
  
  meanVec = NULL # creating placeholders
  sdVec = NULL
  countVec = NULL
  
  for (i in names(CensusTibble)){ # loop through all the census tibble variable names
    if (i %in% numVars) { # if the name matches one of the specified numeric variables... 
      CensusDropNA <- CensusTibble |> filter(!is.na(CensusTibble[[i]])) # create temporary CensusDropNA which removes the NA rows of the i variable 
      meanVec[[i]] <- sum(CensusDropNA[[i]]*CensusDropNA[["PWGTP"]])/sum(CensusDropNA[["PWGTP"]]) # build a vector of calculated weighted means, naming each element per the index
      sdVec[[i]] <- sqrt(sum(CensusDropNA[[i]]^2*CensusDropNA[["PWGTP"]])/sum(CensusDropNA[["PWGTP"]])-meanVec[[i]]^2) # build a vector of calculated weighted sd, naming each element per the index
    }
    if (i %in% catVars) { # if the name matches one of the specified categorical variables...
      countVec[[i]] <- table(CensusTibble[[i]]) # use table() to return un-weighted counts for the levels of the factors
    }
  }
  CombinedSummary = list("mean" = meanVec, "sd" = sdVec, "count" = countVec) # combine the three summary named vectors into a list
  return(CombinedSummary)
} 
```

Now some tests of `summary()`

```{r}
# testing all variables across multiple years
summary(QMultiCensus(CensusYears = c(2015:2017), numVars = c("PWGTP", "AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP"), catVars = c("FER", "HHL", "SCH", "SCHL", "SEX"), geoLevel = "state", geoSubset = "02"))
```

```{r}
# testing select variables AGEP and HHL for one year
summary(QCensus(year = 2018, numVars = c("AGEP", "JWDP", "JWAP", "GRPIP"), catVars = c("FER", "HHL"), geoLevel = "division", geoSubset = "4" ), numVars = c("AGEP"), catVars = c("HHL"))
```

All tests of `summary()` are successful.

## 9. `plot()` for Census Data

We will write our own function to direct how `plot()` is applied to "census" objects

```{r}
plot.census <- function(CensusTibble, numVar = "AGEP" , catVar = "SEX"){  # setting these AGEP and SEX as the defaults
  
  ValidNumVars <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP") # list of all valid variables for numVar
  ValidCatVars <- c("FER", "HHL", "SCH", "SCHL", "SEX") # list of all valid variables for catVar
 
  #checking if the requested variables are valid
  if (!(numVar %in% ValidNumVars)) {stop("Invalid numeric variable.  Valid options are AGEP, GASP, GRPIP, JWAP, JWDP, JWMNP")} 
  if (!(catVar %in% ValidCatVars)) {stop("Invalid categorical variable.  Valid options are FER, HHL, SCH, SCHL, SEX")} 
  
  #checking if the requested variables exist in the tibble
  if (!(numVar %in% names(CensusTibble))) {stop(paste0(numVar, " is not found in tibble"))}
  if (!(catVar %in% names(CensusTibble))) {stop(paste0(catVar, " is not found in tibble"))}
  
  g <- ggplot(CensusTibble |> drop_na(all_of(numVar), all_of(catVar))) # build plot that drops rows that have NA for either variable (not sure why all_of() is required instead of get())
  g + 
    geom_boxplot(aes(x = get(catVar), y = get(numVar), weight = PWGTP, fill = get(catVar))) + # set up boxplot and apply weights
    labs(x = catVar, y = numVar, title = paste0(numVar, " by ", catVar)) + # set x label, y label, and title
    scale_fill_discrete(catVar) + # set legent label
    theme(plot.title = element_text(hjust = 0.5)) # center the title

}
```

Now some tests of `plot()`

```{r}
plot(QCensus(numVars = "AGEP", catVars = "SCH"), numVar = "AGEP", catVar = "SCH")
```

```{r}
plot(QMultiCensus(CensusYears = c(2021:2022), numVars = "GASP", catVars = "HHL", geoSubset = "02" ), numVar = "GASP", catVar = "HHL")
```

All tests of `plot()` are successful.

## 10. Data Investigation

Let's explore the relationship between Job Arrival Time (JWAP) and School Enrollment (SCH).

To begin, we will use `QMultiCensus()` to grab census data from 2010 to 2022 (excluding 2020). We will consider both Rhode Island (a predominantly urban state) and South Dakota (a heavy farming state) to see if any contrast is apparent in the data.

```{r}
RIData <- QMultiCensus(CensusYears = 2010:2022, numVars = "JWAP", catVars = "SCH", geoLevel = "state", geoSubset = "44") # 44 is Rhode Island
```

```{r}
SDData <- QMultiCensus(CensusYears = 2010:2022, numVars = "JWAP", catVars = "SCH", geoLevel = "state", geoSubset = "46") # 46 is South Dakota
```

Next we use `plot()` on our census data to create box plots. Titles, labels, and subtitles will be amended to be more descriptive of the dataset and specific variable details.

```{r}
#| warning: false # hiding warnings/messages related to overwrite of the labels
#| message: false
plot(RIData, numVar = "JWAP", catVar = "SCH") +
  labs(x = "School Enrollment", y = "Job Arrival Time", title = "Job Arrival Time by School Enrollment", subtitle = "Rhode Island 2010-2022 (excluding 2020)") +
  scale_fill_discrete(name = "School Enrollment", labels = c("1" = "No w/in 3 mo", "2" = "Yes, Public", "3" = "Yes, Private")) + # modifying factor level labels
  scale_x_discrete(name = "School Enrollment") +
  theme(plot.subtitle = element_text(hjust = 0.5)) # Centering subtitle
```

```{r}
#| warning: false # hiding warnings/messages related to overwrite of the labels
#| message: false
plot(SDData, numVar = "JWAP", catVar = "SCH") +
  labs(x = "School Enrollment", y = "Job Arrival Time", title = "Job Arrival Time by School Enrollment", subtitle = "South Dakota 2010-2022 (excluding 2020)") +
  scale_fill_discrete(name = "School Enrollment", labels = c("1" = "No w/in 3 mo", "2" = "Yes, Public", "3" = "Yes, Private")) +
  scale_x_discrete(name = "School Enrollment") +
  theme(plot.subtitle = element_text(hjust = 0.5))
```

Review of the RI and SD boxplots provide the following observations:

1.  Job arrival time is later in the day for current students vs non-students

2.  Job arrival time has more variation for current students vs non-students

3.  Job arrival time among non-students is earlier in the day for SD compared to RI

Let's reinforce these plot observations with calculated statistics. We will use `summary()` on our census data to calculate means and standard deviations.

```{r}
# For #1 above, grab the mean JWAP (both states combined) for students vs non-students
print(CombinedMeanStudents <- summary(filter(bind_rows(RIData, SDData), SCH %in% c("2","3")), numVars = "JWAP", catVars = "SCH")$mean$JWAP)
print(CombinedMeanNonStudents <- summary(filter(bind_rows(RIData, SDData), SCH == "1"), numVars = "JWAP", catVars = "SCH")$mean$JWAP)
```

```{r}
# For #2 above, grab the sd JWAP (both states combined) for students vs non-students
print(CombinedSDStudents <- summary(filter(bind_rows(RIData, SDData), SCH %in% c("2","3")), numVars = "JWAP", catVars = "SCH")$sd$JWAP)
print(CombinedSDNonStudents <- summary(filter(bind_rows(RIData, SDData), SCH == "1"), numVars = "JWAP", catVars = "SCH")$sd$JWAP)
```

```{r}
# For #3 above, grab the mean and sd JWAP of both RI and SD for non-students
print(RINonStudentsMean <- summary(filter(RIData, SCH == 1), numVars = "JWAP", catVars = "SCH")$mean$JWAP)
print(SDNonStudentsMean <- summary(filter(SDData, SCH == 1), numVars = "JWAP", catVars = "SCH")$mean$JWAP)
print(RINonStudentsSD <- summary(filter(RIData, SCH == 1), numVars = "JWAP", catVars = "SCH")$sd$JWAP)
print(SDNonStudentsSD <- summary(filter(SDData, SCH == 1), numVars = "JWAP", catVars = "SCH")$sd$JWAP)
```

Expanding on the prior observations with statistical results:

The following observations draw on combined PUMS data for Rhode Island and South Dakota across survey years 2010 through 2022 (excluding 2020 due to the pandemic). Observations 1 and 2 pool both states. Observation 3 contrasts them.

1.  Current students arrive at work later in the day than non-students. The weighted mean arrival time is 10:41 a.m. (`r round(CombinedMeanStudents)` minutes after midnight) for current students, compared to 8:36 a.m. (`r round(CombinedMeanNonStudents)` minutes) for non-students.  This represents a difference of more than two hours. A plausible explanation is that many students work after-school jobs, which start later in the day.

2.  Arrival time is more variable for current students than non-students. The weighted standard deviation is `r round(CombinedSDStudents)` minutes for students versus `r round(CombinedSDNonStudents)` minutes for non-students. This may reflect students working irregular part-time hours scheduled around their classes, producing a wider spread of start times.

3.  Among non-students, arrival time is earlier in South Dakota than in Rhode Island. The weighted mean is 8:23 a.m. (`r round(SDNonStudentsMean)` minutes) in South Dakota versus 8:47 a.m. (`r round(RINonStudentsMean)` minutes) in Rhode Island, with standard deviations of `r round(SDNonStudentsSD)` and `r round(RINonStudentsSD)` minutes respectively. One possible explanation is a difference in the dominant job types between the two states: South Dakota is largely rural/agricultural and farm work tends to start early. Rhode Island is more urban and may skew toward office and service jobs with later start times. 

Evaluation of statistical significance of the above observed differences is recommended, though beyond the scope of this analysis.
