---
title: "Project1"
author: "Ryan Friedman"
format: html
editor: visual
---

```{r}
#| include: false
# load packages that may be used
library(tidyverse)
library(readxl)
library(DBI)
library(knitr)
library(httr)
library(jsonlite)
library(psych)
library(gt)
library(tidycensus)
```

```{r}
#| include: false
APIKey <- "5a644b1ecc301cab1403e0940017740912735ef9" # R.Friedman API Key for census.gov
```

## Test Query of API

To start, we'll do a test query of the API and review the format of the data

```{r}
TestData <- GET(paste0("https://api.census.gov/data/2022/acs/acs1/pums?get=SEX,PWGTP,AGEP&for=state:06&SCHL=24&key=", APIKey))
```

```{r}
TestParsed <- fromJSON(rawToChar(TestData$content))
head(TestParsed)
```

Looks like we get a matrix of all characters. Also the first row contains the variable names. We need to:

1.  Set the names based on the first row

2.  Delete the first row

3.  Convert to tibble

4.  Convert variables into numeric or factors

```{r}
TestCleaned <- TestParsed
colnames(TestCleaned) <- TestCleaned[1, ] # set the column names from the first row
TestCleaned <- as_tibble(TestCleaned[-1, ]) # delete first row and convert to tibble
TestCleaned <- TestCleaned |>
  mutate(SEX = factor(SEX), PWGTP = as.numeric(PWGTP), AGEP = as.numeric(AGEP), SCHL = factor(SCHL), state = factor(state)) # setting variables as numeric or factor
head(TestCleaned)
str(TestCleaned)
```

## Helper Function `IntoTibble()`

`IntoTibble()` function is designed to take the census data returned from the `GET()` and convert into a tibble similar to the steps above

```{r}
IntoTibble <- function (CensusGet){ # input is the get() data from the census API
  CensusParsed <- fromJSON(rawToChar(CensusGet$content)) # extracts the content from the API get
  colnames(CensusParsed) <- CensusParsed[1, ] # sets variable names based on the first row
  CensusParsed <- as_tibble(CensusParsed[-1, ]) # deletes first row and converts to tibble
  CensusParsed
}

head(IntoTibble(TestData)) # testing against the above data
```

Test of `IntoTibble()` function is successful.

## Single Year Query Function `QCensus()`

`QCensus()` will query the API and allow user to specify:

- Year of survey

- Numeric variables (AGEP, GASP, GRPIP, JWAP, JWDP, JWMNP)

  - AGEP and PWGTP are default

  - PWGTP is always returned

- Categorical variables (FER, HHL, SCH, SCHL, SEX)

  - SEX is default

- Geography level (region, division, state)

- Optional geography subsetting within the API call

```{r}
QCensus <- function (year = 2022,
                     numVars = c("AGEP"), 
                     catVars = c("SEX"), 
                     geoLevel = "state",
                     geoSubset = "56", # setting Wyoming as default state
                     APIKey = "5a644b1ecc301cab1403e0940017740912735ef9") # using R.Friedman API key as default
  { 
  
  if (geoSubset == "all") {geoSubset = "*"} # converts entry of "all" into "*" 
  
  if (!is.numeric(year)){stop("Year must be a number")} # first checks if year is a number
  if (year == 2020 | year %% 1 != 0 | year > 2022 | year < 2010){ # check if year is valid integer between 2010 and 2022, excluding 2020
    stop("Invalid year.  Must be an integer from 2010 to 2022, excluding 2020")}
  
  ValidNumVars <- c("PWGTP", "AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP") # list of all valid variables for numVars
  ValidCatVars <- c("FER", "HHL", "SCH", "SCHL", "SEX") # list of all valid variables for catVars
  ValidLevelVars <- c("region", "division", "state") # list of all valid variables for geoLevel
  
  if (!all(numVars %in% ValidNumVars)) { # confirm numeric variables are valid
    stop("Invalid numeric variable(s).  Valid numeric variables are: PWGTP, AGEP, GASP, GRPIP, JWAP, JWDP, JWMNP.")}  
  if (!all(catVars %in% ValidCatVars)) { # confirm categorical variables are valid
    stop("Invalid categorical variable(s).  Valid categorical variables are: FER, HHL, SCH, SCHL, SEX.")}    
  if (!all(geoLevel %in% ValidLevelVars)) { # confirm geography level variables are valid
    stop("Invalid geography level variable(s).  Valid geography levels are: region, division, state.")}    
  
  if (length(unique(numVars)) != length(numVars)) { # checking for duplicated numeric variables
    stop("Error: Duplicated numeric variable")}
  if (length(unique(catVars)) != length(catVars))  { # checking for duplicated categorical variables
    stop("Error: Duplicated categorical variable")}
  
  numVars <- numVars[numVars != "PWGTP"] # pulling out PWGTP from numVars since that variable is always included anyways
  
  if(length(numVars) < 1 ) { # checks minimum input requirement for numVars
    stop("Must specify at least one numeric variable other than PWGTP")}  
  if(length(catVars) < 1 ) { # checks minimum input requirement for catVars
    stop("Must specify at least one categorical variable")} 
  
  # establish the valid codes for regions, divisions, and states
  ValidRegions <- c("1", "2", "3", "4", "*")
  ValidDivisions <- c("1", "2", "3", "4", "5", "6", "7", "8", "9", "*")
  ValidStates <- c("01", "02", "04", "05", "06", "08", "09", "10", "11", "12", "13", "15", "16", "17", "18", "19", "20", 
                   "21", "22", "23", "24", "25", "26", "27", "28", "29", "30", "31", "32", "33", "34", "35", "36", "37", 
                   "38", "39", "40", "41", "42", "44", "45", "46", "47", "48", "49", "50", "51", "53", "54", "55", "56", "*")
  
  # add corresponding names (to be used later).  "All" is simply a placeholder that will not be used
  names(ValidDivisions) <- c("New England", "Middle Atlantic", "East North Central", "West North Central", 
                             "South Atlantic", "East South Central", "West South Central", "Mountain", "Pacific", "All")
  names(ValidRegions) <- c("Northeast", "Midwest", "South", "West", "All")
  names(ValidStates) <- c("Alabama", "Alaska", "Arizona", "Arkansas", "California", "Colorado", "Connecticut", "Delaware", 
                          "District of Columbia", "Florida", "Georgia", "Hawaii", "Idaho", "Illinois", "Indiana", "Iowa", 
                          "Kansas", "Kentucky", "Louisiana", "Maine", "Maryland", "Massachusetts", "Michigan", "Minnesota", 
                          "Mississippi", "Missouri", "Montana", "Nebraska", "Nevada", "New Hampshire", "New Jersey", 
                          "New Mexico", "New York", "North Carolina", "North Dakota", "Ohio", "Oklahoma", "Oregon", 
                          "Pennsylvania", "Rhode Island", "South Carolina", "South Dakota", "Tennessee", "Texas", "Utah", 
                          "Vermont", "Virginia", "Washington", "West Virginia", "Wisconsin", "Wyoming", "All")
  
  if (geoLevel == "region" & !(geoSubset %in% ValidRegions)) { # check if geoSubset matches valid regions
    stop("Invalid region code")} 
  if (geoLevel == "division" & !(geoSubset %in% ValidDivisions)) { # check if geoSubset matches valid divisions
    stop("Invalid division code")} 
  if (geoLevel == "state" & !(geoSubset %in% ValidStates)) { # check if geoSubset matches valid states
    stop("Invalid state code")} 
  
  
  APIURL <- paste0("https://api.census.gov/data/", year, "/acs/acs1/pums?get=PWGTP,") # begins the URL including year and PWGTP
  APIURL <- paste0(APIURL, paste(numVars, collapse = ","), ",", paste(catVars, collapse = ",")) # adds numeric variables and categorical variables
  APIURL <- paste0(APIURL, "&for=", geoLevel, ":", geoSubset) # adds geoLevel and subset
  APIURL <- paste0(APIURL, "&key=", APIKey) # adds API key
  
  QCensusParsed <- IntoTibble(GET(APIURL))
  
  # check API return success?
  # fix variable types
  
  return(QCensusParsed)
  
  
}
```

## Multi-Year Query Function `QMultiCensus()`
